\bf The forestecology R package for fitting and assessing models of interspecies competitive effects on the growth of trees

### 2021-02-25 paper/forestecology.tex

\bigskip
\begin{abstract}
1. Many models for the growth of trees that incorporate the effect of
interspecies competition are based on a neighborhood effect assumption
whereby all trees within a fixed distance of all focal trees are
considered competitors. Methods and tools are needed to quantify this
competitive effect and assess the quality of all resulting models\\
2. We present the \texttt{forestecology} package providing methods for
both 1) evaluating the effect of competitor species identity using
permutation tests and 2) evaluating model performance using spatial
cross-validation. Following \citet{allen_permutation_2020}, we implement
a Bayesian linear regression competition model.\\
3. We demonstrate the package's functionality using data from the
Smithsonian Conservation Biology Institute's large forest dynamics plot,
part of the ForestGEO global network of reseach sites. Given ForestGEO's
data collection protocols and data formatting standards, the package was
designed with cross-site compatibility in mind. We demonstrate that both
1) competitor species identity matters and 2) that not spatially
cross-validating leads to error estimates that are overly optimistic.\\
4. The package features 1) \texttt{tidyverse}-like structure whereby
verb-named functions can be modularly ``piped'' in sequence, 2)
functions with standardized inputs/outputs of simple features
\texttt{sf} package class, and 3) R S3 object-oriented implementation of
the Bayesian linear regression model. These three facts allow for clear
articulation of all the steps in the sequence of analysis and easy
wrangling and visualization of the geospatial forestry data.
Furthermore, while the package only has Bayesian linear regression
implemented, the package was designed with extensibility to other
methods in mind.
\end{abstract}

@@

\[
\newcommand{\dbh}{\text{DBH}}
\newcommand{\biomass}{\text{biomass}}
\newcommand{\BA}{\text{BA}}
y_{ij} = f(\vec{x}_{ij}) + \epsilon_{ij} = \beta_{0,j} + \beta_{\dbh,j} \cdot \dbh_{ij} + \sum_{k=1}^{K} \lambda_{jk} \cdot \BA_{ijk} + \epsilon_{ij}
\]

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{growth_scbi <-}
\StringTok{  }\KeywordTok{compute_growth}\NormalTok{(}
    \DataTypeTok{census_1 =}\NormalTok{ census_}\DecValTok{2013}\NormalTok{_scbi,}
    \DataTypeTok{census_2 =}\NormalTok{ census_}\DecValTok{2018}\NormalTok{_scbi }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(}\OperatorTok{!}\KeywordTok{str_detect}\NormalTok{(codes, }\StringTok{"R"}\NormalTok{)),}
    \DataTypeTok{id =} \StringTok{"stemID"}
\NormalTok{  )}
\NormalTok{growth_scbi}
\CommentTok{## Simple feature collection with 7954 features and 8 fields}
\CommentTok{## geometry type:  POINT}
\CommentTok{## dimension:      XY}
\CommentTok{## bbox:           xmin: 0.2 ymin: 300 xmax: 299.9 ymax: 600}
\CommentTok{## CRS:            NA}
\CommentTok{## # A tibble: 7,954 x 9}
\CommentTok{##    stemID sp     dbh1 codes1 status  dbh2 codes2  growth     geometry}
\CommentTok{##     <dbl> <fct> <dbl> <chr>  <chr>  <dbl> <chr>    <dbl>      <POINT>}
\CommentTok{##  1      4 nysy  13.6  M      A       14.2 M       0.103  (14.2 428.5)}
\CommentTok{##  2      5 havi   8.8  M      A        9.6 M;P     0.150   (9.4 436.4)}
\CommentTok{##  3      6 havi   3.25 NULL   A        4   M       0.140     (1.3 434)}
\CommentTok{##  4     77 qual  65.2  M      A       66   M       0.141  (34.7 307.2)}
\CommentTok{##  5     79 tiam  47.7  M      A       46.8 M      -0.161    (40 381.1)}
\CommentTok{##  6     80 caca   5.15 M      A        6.5 M       0.253  (38.7 421.7)}
\CommentTok{##  7     96 libe   2.3  J;M    A        3.7 M       0.262      (60 310)}
\CommentTok{##  8    100 caca   5.09 NULL   A       NA   DN     NA      (52.5 476.3)}
\CommentTok{##  9    101 litu  65.4  M      A       68.4 M       0.552  (47.1 567.3)}
\CommentTok{## 10    102 astr   1.99 NULL   A        2.5 M       0.0954 (40.8 575.5)}
\CommentTok{## # ... with 7,944 more rows}
\end{Highlighting}
\end{Shaded}

The output \texttt{growth\_scbi} is a single data frame of class
\texttt{sf} that includes \texttt{growth}, the average annual growth in
DBH (cm y\textsuperscript{-1}) for all stems that were alive at both
time points, as well a \texttt{geometry} variable encoding each tree's
geolocation. Furthermore, the variables that should remain unchanged
between censuses appear only once, such as location variables
\texttt{gx} and \texttt{gy}; as well as species-related variables.
Variables that should change between censuses are suffixed with
\texttt{1} and \texttt{2} indicating the earlier and later censuses,
such as \texttt{dbh1/dbh2} and \texttt{codes1/codes2}.

The data format of other sites may be such that our
\texttt{compute\_growth()} function doesn't work at all. However, in the
end all that matters is that the growth of all stems is saved in a data
frame of class \texttt{sf} whereby the geolocation of each tree is
presented in a \texttt{geometry} variable of type
\texttt{\textless{}POINT\textgreater{}} and at a minimum the data
contains the variables above (DAVE NOTE: THis isn't exaclty clear to me
what variables above means. Maybe list all the variables needed. Do they
have to have exact names?)

@@

The second element of spatial information are blocks corresponding to
folds of a spatial cross-validation algorithm used to estimate
out-of-sample model error. Conventional cross-validation algorithms
assign observations to folds by randomly resampling individual
observations. However, an assumption that the observations are
independent underlies any cross-validation algorithm. In the case of
forest census data, observations exhibit spatial autocorrelation. We
therefore incorporate this spatial dependence into the cross-validation
algorithm by randomly resampling spatial blocks of trees
\citet{roberts_cross-validation_2017}
\citet{pohjankukka_estimating_2017}. (DAVE NOTE: I don't want to change
somethign I don't understand, but ``randomly resampling'' seems wrong to
me. I don't see anything random about our block structure. Seems to me
like ``incorpoate this spatial dependence into the cross-validation
algorithm with our spatial blocks of trees.'')

@@

DAVE NOTE::there are a couple of places where you convert to factor.
Earlier with \texttt{sp} and then twice in the code block below. ANy
reason we can't just do this ``under the hood'' in the package code?
Also any reason with swtiched from undercase with underscores to
\texttt{SpatialBlock\_scbi} in the naming?

@@

\CommentTok{# Associate each observation to a fold}
\NormalTok{SpatialBlock_scbi <-}\StringTok{ }\KeywordTok{spatialBlock}\NormalTok{(}
  \DataTypeTok{speciesData =}\NormalTok{ growth_scbi, }\DataTypeTok{k =}\NormalTok{ n_fold, }\DataTypeTok{selection =} \StringTok{"systematic"}\NormalTok{, }
  \DataTypeTok{blocks =}\NormalTok{ blocks_scbi, }\DataTypeTok{showBlocks =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{verbose =} \OtherTok{FALSE}
\NormalTok{)}

\NormalTok{growth_scbi <-}\StringTok{ }\NormalTok{growth_scbi }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{foldID =}\NormalTok{ SpatialBlock_scbi}\OperatorTok{$}\NormalTok{foldID }\OperatorTok{%>%}\StringTok{ }\KeywordTok{factor}\NormalTok{())}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# }\AlertTok{TODO}\CommentTok{: Below reconcile the number of rows as they off by one from growth_scbi %>% }
\CommentTok{# filter(!is.na(growth) & !buffer). Perhaps by removing NA's in the growth_scbi stage?}
\NormalTok{focal_vs_comp_scbi <-}\StringTok{ }\NormalTok{growth_scbi }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{create_focal_vs_comp}\NormalTok{(comp_dist, }\DataTypeTok{cv_grid_sf =}\NormalTok{ blocks_scbi, }\DataTypeTok{id =} \StringTok{"stemID"}\NormalTok{)}
\NormalTok{focal_vs_comp_scbi }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{select}\NormalTok{(}\OperatorTok{-}\NormalTok{foldID)}
\CommentTok{## # A tibble: 6,296 x 6}
\CommentTok{##    focal_ID focal_sp   dbh     geometry  growth comp             }
\CommentTok{##       <dbl> <fct>    <dbl>      <POINT>   <dbl> <list>           }
\CommentTok{##  1        4 nysy     13.6  (14.2 428.5)  0.103  <tibble [20 x 4]>}
\CommentTok{##  2        5 havi      8.8   (9.4 436.4)  0.150  <tibble [32 x 4]>}
\CommentTok{##  3       79 tiam     47.7    (40 381.1) -0.161  <tibble [20 x 4]>}
\CommentTok{##  4       80 caca      5.15 (38.7 421.7)  0.253  <tibble [12 x 4]>}
\CommentTok{##  5       96 libe      2.3      (60 310)  0.262  <tibble [14 x 4]>}
\CommentTok{##  6      101 litu     65.4  (47.1 567.3)  0.552  <tibble [19 x 4]>}
\CommentTok{##  7      102 astr      1.99 (40.8 575.5)  0.0954 <tibble [44 x 4]>}
\CommentTok{##  8      126 cato     37.4  (60.6 400.2)  0.165  <tibble [16 x 4]>}
\CommentTok{##  9      127 caca      8.72 (72.7 514.1)  0.0370 <tibble [14 x 4]>}
\CommentTok{## 10      139 astr      1.71 (96.7 315.1)  0.0549 <tibble [48 x 4]>}
\CommentTok{## # ... with 6,286 more rows}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{focal_vs_comp_scbi}
\CommentTok{## # A tibble: 6,296 x 8}
\CommentTok{##    focal_ID focal_sp   dbh foldID                  geometry  growth}
\CommentTok{##       <dbl> <fct>    <dbl> <fct>                    <POINT>   <dbl>}
\CommentTok{##  1        4 nysy     13.6  1                   (14.2 428.5)  0.103 }
\CommentTok{##  2        5 havi      8.8  1                    (9.4 436.4)  0.150 }
\CommentTok{##  3       79 tiam     47.7  1                     (40 381.1) -0.161 }
\CommentTok{##  4       80 caca      5.15 1                   (38.7 421.7)  0.253 }
\CommentTok{##  5       96 libe      2.3  1                       (60 310)  0.262 }
\CommentTok{##  6      101 litu     65.4  3                   (47.1 567.3)  0.552 }
\CommentTok{##  7      102 astr      1.99 3                   (40.8 575.5)  0.0954}
\CommentTok{##  8      126 cato     37.4  1                   (60.6 400.2)  0.165 }
\CommentTok{##  9      127 caca      8.72 3                   (72.7 514.1)  0.0370}
\CommentTok{## 10      139 astr      1.71 1                   (96.7 315.1)  0.0549}
\CommentTok{## # ... with 6,286 more rows, and 2 more variables: comp <list>,}
\CommentTok{## #   growth_hat <dbl>}
\end{Highlighting}
\end{Shaded}

We then compare the observed and fitted/predicted growths to compute the
root mean squared error (RMSE) of our model fit. DAVE NOTE: I rounded
all the RMSEs, in ecology convention to not show WAY more digits than
resolution of the measurement.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{model_rmse <-}\StringTok{ }\NormalTok{focal_vs_comp_scbi }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{rmse}\NormalTok{(}\DataTypeTok{truth =}\NormalTok{ growth, }\DataTypeTok{estimate =}\NormalTok{ growth_hat) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{pull}\NormalTok{(.estimate) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{round}\NormalTok{(}\DecValTok{3}\NormalTok{)}
\NormalTok{model_rmse}
\CommentTok{## [1] 0.128}
\end{Highlighting}
\end{Shaded}

@@

These plots give the posterior distributions of parameters from equation
1 (DAVE NOTE:SOrry couldn't figure out how to make equation references
work). For many pacakge users they will be of interest because they give
insight into the species-specific competitive interactions. Setting
\texttt{type\ =\ "intercepts"} gives posterior distributions for
\(\beta_{0,j}\) and \texttt{type\ =\ "dbh\_slopes"} for
\(\beta_{dbh,j}\). These give species specific growth independent of
competition. The values of more interest are plotted with
\texttt{type\ =\ \textquotesingle{}competition\textquotesingle{}} which
gives the posterior distribution for \(\lambda_{j,k}\) species-specific
competition coefficients (i.e., the \(\lambda\)-matrix). Negative values
indicate a competitor species which slows the growth of a focal species.
Here, for example, we see that tulip trees (\texttt{litu}) have a strong
negative effect on the growth of conspecifics but relatively little
effect on neighbors of the other two species.

@@

\NormalTok{focal_vs_comp_scbi <-}\StringTok{ }\NormalTok{focal_vs_comp_scbi }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{growth_hat_shuffle =} \KeywordTok{predict}\NormalTok{(comp_bayes_lm_scbi_shuffle, }\DataTypeTok{newdata =}\NormalTok{ focal_vs_comp_scbi))}

\NormalTok{model_rmse_shuffle <-}\StringTok{ }\NormalTok{focal_vs_comp_scbi }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{rmse}\NormalTok{(}\DataTypeTok{truth =}\NormalTok{ growth, }\DataTypeTok{estimate =}\NormalTok{ growth_hat_shuffle) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{pull}\NormalTok{(.estimate) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{round}\NormalTok{(}\DecValTok{3}\NormalTok{)}
\NormalTok{model_rmse_shuffle}
\CommentTok{## [1] 0.131}
\end{Highlighting}
\end{Shaded}

@@

To this end, we use the \texttt{foldID} variable defined in Section
\ref{spatial-information} whereby all focal trees are assigned to one of
4 spatially contiguous blocks that act as folds in our cross-validation
routine. Figure \ref{fig:scbi-spatial-cross-validation-schematic}
presents a schematic illustrating this scheme for fold 1 (bottom-left)
as the test set and folds 2, 3, and 4 as the training sets. We fit the
model to all focal trees in the training set, apply the model to all
focal trees in the test set to compute fitted/predicted values, and
compute the RMSE of the observed versus predicted growths. We repeat
this procedure 3 more times with each of the three remaining folds
acting as the test set and then average all four resulting RMSE's.
Furthermore, in order to maintain spatial independence between the test
and training set, a fold buffer that extend outwards from the boundary
of the test set is computed; all trees falling within this fold buffer
are excluded from the training set. DAVE NOTE: I repalced ``buffer''
with ``fold buffer'' just to keep clear that this is different than our
global buffer. But if you think that distinction is not necessary feel
free to switch back!

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{model_rmse_cv <-}\StringTok{ }\NormalTok{focal_vs_comp_scbi }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{rmse}\NormalTok{(}\DataTypeTok{truth =}\NormalTok{ growth, }\DataTypeTok{estimate =}\NormalTok{ growth_hat) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{pull}\NormalTok{(.estimate) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{round}\NormalTok{(}\DecValTok{3}\NormalTok{)}
\NormalTok{model_rmse_cv}
\CommentTok{## [1] 0.14}
\end{Highlighting}
\end{Shaded}

>>>>

\bigskip
\begin{abstract}
1. Many models for the growth of trees that incorporate interspecies
competition are based on a neighborhood effect assumption whereby all
trees within a fixed distance of a focal tree are considered
competitors. Methods and tools are needed to quantify this competitive
effect and assess the quality of all resulting models.\\
2. We present the \texttt{forestecology} package providing methods for
both 1) evaluating the effect of competitor species identity using
permutation tests and 2) evaluating model performance using spatial
cross-validation. Following \citet{allen_permutation_2020}, we implement
a Bayesian linear regression competition model.\\
3. We demonstrate the package's functionality using data from the
Smithsonian Conservation Biology Institute's large forest dynamics plot,
part of the ForestGEO global network of reseach sites. Given ForestGEO's
data collection protocols and data formatting standards, the package was
designed with cross-site compatibility in mind. We demonstrate that both
1) competitor species identity matters and 2) that not spatially
cross-validating leads to error estimates that are overly optimistic.\\
4. The package features 1) \texttt{tidyverse}-like structure whereby
verb-named functions can be modularly ``piped'' in sequence, 2)
functions with standardized inputs/outputs of simple features
\texttt{sf} package class, and 3) an S3 object-oriented implementation
of the Bayesian linear regression model. These three facts allow for
clear articulation of all the steps in the sequence of analysis and easy
wrangling and visualization of the geospatial forestry data.
Furthermore, while the package only has Bayesian linear regression
implemented, the package was designed with extensibility to other
methods in mind.
\end{abstract}

@@

\begin{equation}
\label{eq:model}
y_{ij} = f(\vec{x}_{ij}) + \epsilon_{ij} = \beta_{0,j} + \beta_{\text{dbh},j} \cdot \text{dbh}_{ij} + \sum_{k=1}^{K} \lambda_{jk} \cdot \text{BA}_{ijk} + \epsilon_{ij}
\end{equation}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{growth_scbi <-}
\StringTok{  }\KeywordTok{compute_growth}\NormalTok{(}
    \DataTypeTok{census_1 =}\NormalTok{ census_}\DecValTok{2013}\NormalTok{_scbi,}
    \DataTypeTok{census_2 =}\NormalTok{ census_}\DecValTok{2018}\NormalTok{_scbi }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(}\OperatorTok{!}\KeywordTok{str_detect}\NormalTok{(codes, }\StringTok{"R"}\NormalTok{)),}
    \DataTypeTok{id =} \StringTok{"stemID"}
\NormalTok{  )}
\NormalTok{growth_scbi}
\CommentTok{## Simple feature collection with 7954 features and 8 fields}
\CommentTok{## geometry type:  POINT}
\CommentTok{## dimension:      XY}
\CommentTok{## bbox:           xmin: 0.2 ymin: 300 xmax: 300 ymax: 600}
\CommentTok{## CRS:            NA}
\CommentTok{## # A tibble: 7,954 x 9}
\CommentTok{##   stemID sp     dbh1 codes1 status  dbh2 codes2 growth   geometry}
\CommentTok{##    <dbl> <fct> <dbl> <chr>  <chr>  <dbl> <chr>   <dbl>    <POINT>}
\CommentTok{## 1      4 nysy  13.6  M      A       14.2 M       0.103 (14.2 428)}
\CommentTok{## 2      5 havi   8.8  M      A        9.6 M;P     0.150  (9.4 436)}
\CommentTok{## 3      6 havi   3.25 NULL   A        4   M       0.140  (1.3 434)}
\CommentTok{## 4     77 qual  65.2  M      A       66   M       0.141 (34.7 307)}
\CommentTok{## 5     79 tiam  47.7  M      A       46.8 M      -0.161   (40 381)}
\CommentTok{## # ... with 7,949 more rows}
\end{Highlighting}
\end{Shaded}

The output \texttt{growth\_scbi} is a single data frame of class
\texttt{sf} that includes variables \texttt{growth}, the average annual
growth in DBH (cm y\textsuperscript{-1}) for all stems that were alive
at both time points, and \texttt{geometry}, the \texttt{sf} package's
encoding of geolocations of type
\texttt{\textless{}POINT\textgreater{}}. Furthermore, the variables that
should remain unchanged between censuses appear only once, such as
location variables \texttt{gx} and \texttt{gy}; as well as
species-related variables. Variables that should change between censuses
are suffixed with \texttt{1} and \texttt{2} indicating the earlier and
later censuses, such as \texttt{dbh1/dbh2} and \texttt{codes1/codes2}.
The data format of other sites may be such that our
\texttt{compute\_growth()} function doesn't work at all. However, in the
end all that matters is that the growth of all stems is saved in a data
frame of class \texttt{sf} and at a minimum the data contains the
variables above. TODO: enumerate what these variables are and say if
they need exact names

@@

The second element of spatial information are blocks corresponding to
folds of a spatial cross-validation algorithm used to estimate
out-of-sample model error. Conventional cross-validation algorithms
assign observations to folds by randomly resampling individual
observations. However, an assumption that the observations are
independent underlies any cross-validation algorithm. In the case of
forest census data, observations exhibit spatial autocorrelation. We
therefore incorporate this spatial dependence into the cross-validation
algorithm with our spatial blocks of trees.
\citet{roberts_cross-validation_2017}
\citet{pohjankukka_estimating_2017}.

@@

@@

\CommentTok{# Associate each observation to a fold}
\NormalTok{spatial_block_scbi <-}\StringTok{ }\KeywordTok{spatialBlock}\NormalTok{(}
  \DataTypeTok{speciesData =}\NormalTok{ growth_scbi, }\DataTypeTok{k =}\NormalTok{ n_fold, }\DataTypeTok{selection =} \StringTok{"systematic"}\NormalTok{, }
  \DataTypeTok{blocks =}\NormalTok{ blocks_scbi, }\DataTypeTok{showBlocks =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{verbose =} \OtherTok{FALSE}
\NormalTok{)}

\NormalTok{growth_scbi <-}\StringTok{ }\NormalTok{growth_scbi }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{foldID =}\NormalTok{ spatial_block_scbi}\OperatorTok{$}\NormalTok{foldID }\OperatorTok{%>%}\StringTok{ }\KeywordTok{factor}\NormalTok{())}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# }\AlertTok{TODO}\CommentTok{: Below reconcile the number of rows as they off by one from growth_scbi %>% }
\CommentTok{# filter(!is.na(growth) & !buffer). Perhaps by removing NA's in the growth_scbi stage?}
\NormalTok{focal_vs_comp_scbi <-}\StringTok{ }\NormalTok{growth_scbi }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{create_focal_vs_comp}\NormalTok{(comp_dist, }\DataTypeTok{cv_grid_sf =}\NormalTok{ blocks_scbi, }\DataTypeTok{id =} \StringTok{"stemID"}\NormalTok{)}
\NormalTok{focal_vs_comp_scbi }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{select}\NormalTok{(}\OperatorTok{-}\NormalTok{foldID)}
\CommentTok{## # A tibble: 6,296 x 6}
\CommentTok{##   focal_ID focal_sp   dbh   geometry growth comp             }
\CommentTok{##      <dbl> <fct>    <dbl>    <POINT>  <dbl> <list>           }
\CommentTok{## 1        4 nysy     13.6  (14.2 428)  0.103 <tibble [20 x 4]>}
\CommentTok{## 2        5 havi      8.8   (9.4 436)  0.150 <tibble [32 x 4]>}
\CommentTok{## 3       79 tiam     47.7    (40 381) -0.161 <tibble [20 x 4]>}
\CommentTok{## 4       80 caca      5.15 (38.7 422)  0.253 <tibble [12 x 4]>}
\CommentTok{## 5       96 libe      2.3    (60 310)  0.262 <tibble [14 x 4]>}
\CommentTok{## # ... with 6,291 more rows}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{focal_vs_comp_scbi}
\CommentTok{## # A tibble: 6,296 x 8}
\CommentTok{##   focal_ID focal_sp   dbh foldID                  geometry growth}
\CommentTok{##      <dbl> <fct>    <dbl> <fct>                    <POINT>  <dbl>}
\CommentTok{## 1        4 nysy     13.6  1                     (14.2 428)  0.103}
\CommentTok{## 2        5 havi      8.8  1                      (9.4 436)  0.150}
\CommentTok{## 3       79 tiam     47.7  1                       (40 381) -0.161}
\CommentTok{## 4       80 caca      5.15 1                     (38.7 422)  0.253}
\CommentTok{## 5       96 libe      2.3  1                       (60 310)  0.262}
\CommentTok{## # ... with 6,291 more rows, and 2 more variables: comp <list>,}
\CommentTok{## #   growth_hat <dbl>}
\end{Highlighting}
\end{Shaded}

We then compare the observed and fitted/predicted growths to compute the
root mean squared error (RMSE) of our model fit.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{model_rmse <-}\StringTok{ }\NormalTok{focal_vs_comp_scbi }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{rmse}\NormalTok{(}\DataTypeTok{truth =}\NormalTok{ growth, }\DataTypeTok{estimate =}\NormalTok{ growth_hat) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{pull}\NormalTok{(.estimate)}
\NormalTok{model_rmse}
\CommentTok{## [1] 0.128}
\end{Highlighting}
\end{Shaded}

@@

These plots give the posterior distributions of parameters from Equation
\ref{eq:model}. For many pacakge users they will be of interest because
they give insight into the species-specific competitive interactions.
Setting \texttt{type\ =\ "intercepts"} gives posterior distributions for
\(\beta_{0,j}\) and \texttt{type\ =\ "dbh\_slopes"} for
\(\beta_{dbh,j}\). These give species specific growth independent of
competition. The values of more interest are plotted with
\texttt{type\ =\ \textquotesingle{}competition\textquotesingle{}} which
gives the posterior distribution for \(\lambda_{j,k}\) species-specific
competition coefficients (i.e., the \(\lambda\)-matrix). Negative values
indicate a competitor species which slows the growth of a focal species.
Here, for example, we see that tulip trees (\texttt{litu}) have a strong
negative effect on the growth of conspecifics but relatively little
effect on neighbors of the other two species.

@@

\NormalTok{focal_vs_comp_scbi <-}\StringTok{ }\NormalTok{focal_vs_comp_scbi }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{growth_hat_shuffle =} \KeywordTok{predict}\NormalTok{(comp_bayes_lm_scbi_shuffle, }\DataTypeTok{newdata =}\NormalTok{ focal_vs_comp_scbi))}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{model_rmse_shuffle <-}\StringTok{ }\NormalTok{focal_vs_comp_scbi }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{rmse}\NormalTok{(}\DataTypeTok{truth =}\NormalTok{ growth, }\DataTypeTok{estimate =}\NormalTok{ growth_hat_shuffle) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{pull}\NormalTok{(.estimate)}
\NormalTok{model_rmse_shuffle}
\CommentTok{## [1] 0.131}
\end{Highlighting}
\end{Shaded}

@@

To this end, we use the \texttt{foldID} variable defined in Section
\ref{spatial-information} whereby all focal trees are assigned to one of
4 spatially contiguous blocks that act as folds in our cross-validation
routine. Figure \ref{fig:scbi-spatial-cross-validation-schematic}
presents a schematic illustrating this scheme for fold 1 (bottom-left)
as the test set and folds 2, 3, and 4 as the training sets. We fit the
model to all focal trees in the training set, apply the model to all
focal trees in the test set to compute fitted/predicted values, and
compute the RMSE of the observed versus predicted growths. We repeat
this procedure 3 more times with each of the three remaining folds
acting as the test set and then average all four resulting RMSE's.
Furthermore, in order to maintain spatial independence between the test
and training set, a fold buffer that extend outwards from the boundary
of the test set is computed; all trees falling within this fold buffer
are excluded from the training set.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{model_rmse_cv <-}\StringTok{ }\NormalTok{focal_vs_comp_scbi }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{rmse}\NormalTok{(}\DataTypeTok{truth =}\NormalTok{ growth, }\DataTypeTok{estimate =}\NormalTok{ growth_hat) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{pull}\NormalTok{(.estimate)}
\NormalTok{model_rmse_cv}
\CommentTok{## [1] 0.14}
\end{Highlighting}
\end{Shaded}
